---
layout: post
title: "LR02: SD line, GoA, Regression"
author: "Roberto Bertolusso"
categories: [Linear Regression]
tags: [intubate, magrittr, data science, statistics]
date: "2016-11-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE,
                      fig.align = 'center',
                      fig.width = 6, fig.height = 6.5)
```

This posts continues the discussion of correlation started on <a href="https://rbertolusso.github.io/posts/LR01-correlation">LR01: Correlation</a>.
We will try to answer the following questions:
Should correlation be used for *any* pair of data?
Does association mean causation?
What are ecological correlations?
What happens with the scatter diagram when we change the standard deviations
  of x and y?
What is the SD line?
What is the Graph of averages?
What is the Regression line.
What is the Regression function?

<hr />

<script src="https://cdn.mathjax.org/mathjax/latest/MathJax.js?config=TeX-AMS-MML_HTMLorMML" type="text/javascript"></script>

* Acknowledgments: organization is extracted from:
  - *Freedman, Pisani, Purves, Statistics, 4th ed.*, probably
  the best book on statistical thinking (it maybe has a total of 4-5 formulas). It is referred here as FPP.
  - A lot of what is good is due to Professor Rudy Guerra.

<hr />

### Previous

<a href="https://rbertolusso.github.io/posts/LR01-correlation">LR01: Correlation</a>

### What we know so far

* *Correlation coefficient*: measure of **linear** association,
  or clustering about a line (which line? The 45-degree line?
  The regression line? The SD line? More on this later).

\[
r = \frac{\sum_{i=1}^n(x_i-\bar{x})(y_i-\bar{y})}
    {\sqrt{\sum_{i=1}^n (x_i-\bar{x})^2}\sqrt{\sum_{i=1}^n (y_i-\bar{y})^2}}
\]
  
* *If* the scatterplot of both variables is a *football-shaped* cloud of points,
  we can *summarize* the relationship between both variables by:
    * average of x-values, SD of x-values,
    * average of y-values, SD of y-values.
    * the correlation coefficient r.

```{r, message=FALSE}
library(UsingR)

## Pearson's data
## Father and son data
data(father.son)
x <- father.son$fheight
y <- father.son$sheight
```

```{r}
## Summary statistics
(meanx <- mean(x))
sd(x)

(meany <- mean(y))
sd(y)

(r <- cor(x, y))
```

<br />

### Some situations where correlation should not be used

* Correlation coefficient: useful for *football-shaped* scatter diagrams.

* If not, $r$ can be misleading. Examples that give $r$ almost zero
  but still highly associated:
    - Outliers (that should be rejected **only** if there is good reason to do so)
    - non-linearity.

* Remember: $r$ measures **linear** association, not association in general.

<br />

### Association is not causation

* Example from FPP: "for school children, shoe size is strongly correlated with
  reading skills. However, learning new words does not make the feet get bigger.
  Instead, there is a third factor involved - age. As children get older, they
  learn to read better and they outgrow their shoes. (According to statistical
  jargon (...), age is a **confounder**.) In the example, the confounder was easy to
  spot. Often, this is not so easy. And *the arithmetic of the correlation
  coefficient does not protect you against third factors*."

<br />

### Ecological correlations (misleading use of correlation)
* What are ecological correlations? Correlations done on ecological data? No. It is
  one of (many?) *creative* uses of statistical related techniques
  (another is what is called Simpson's paradox, related to contingency tables,
  and of which you can read on FPP
  or <https://en.wikipedia.org/wiki/Simpson%27s_paradox>,
  or of ecological fallacy, of which you can read on
  <http://www.stat.berkeley.edu/~census/ecofall.txt>).
  
  Link to data:
  <https://rbertolusso.github.io/data/LR/eco_corre.csv>

```{r eco-cor-base, fig.width=8, fig.height=4.5, fig.align='center'}
eco <- read.csv(file = "../data/LR/eco_corre.csv", stringsAsFactors = FALSE)

par(mfrow = c(1,2))

## Left panel
r <- round(cor(eco$math, eco$verbal), 2)
plot(eco$math, eco$verbal, col = eco$group,
     xlim = c(400, 800), ylim = c(400, 800), main = paste0("r1= ", r))

## Right panel
byg <- with(eco, data.frame(math = tapply(math, group, mean),
                            verbal = tapply(verbal, group, mean)))
rg <- round(cor(byg$math, byg$verbal), 2)
plot(byg$math, byg$verbal, col = row.names(byg), pch = 19, cex = 2,
     xlim = c(400, 800), ylim = c(400, 800), main = paste0("r2= ", rg))
```

The same, using `magrittr` pipes, `dplyr` and `intubate` (See
<a href="https://rbertolusso.github.io/posts/intubate-and-stat-functions-in-pipelines">intubate <||> R stat functions in data science pipelines</a> for an introduction
to `intubate`):

```{r eco-cor-intubate, message=FALSE, fig.width=8, fig.height=4.5, fig.align='center'}
library(magrittr)
library(dplyr)
library(intubate)

eco <- read.csv(file = "../data/LR/eco_corre.csv", stringsAsFactors = FALSE)

par(mfrow = c(1,2))

## Left panel
eco %>% ntbt(cor, math, verbal) %>% round(2) -> r
eco %>% 
  ntbt(plot, math, verbal, col = group, main = paste0("r1= ", r),
       xlim = c(400, 800), ylim = c(400, 800))

## Right panel
eco %>%
  group_by(group) %>%
  summarise(math = mean(math),
            verbal = mean(verbal)) -> byg
byg %>% ntbt(cor, math, verbal) %>% round(2) -> rg
byg %>% 
  ntbt(plot, math, verbal, col = group, main = paste0("r2= ", rg),
       xlim = c(400, 800), ylim = c(400, 800), pch = 19, cex = 2)
```


* The left panel contains the raw data, the right panel aggregate data
 (means of each group). Ecological correlation means calculating the
 correlation from the data in the right panel, instead than from data
 in the left panel.
 
* Why do you think someone would like to use the data on the right instead
  of the one on the left?
 
* Superimposing both scatterplots:

```{r eco-cor-both, fig.width=5, fig.height=5.5, fig.align='center'}
eco %>% 
  ntbt(plot, math, verbal, col = group,
       main = paste0("r1= ", r, "; r2= ", rg),
       xlim = c(400, 800), ylim = c(400, 800))
byg %>% 
  ntbt(points, math, verbal, col = group, pch = 19, cex = 2)
```

* Ecological correlations: based on rates or averages (used
 in political science and sociology). They tend to *overstate*
 the strength of an association.

* Correlations based on **rates** or **averages** can be **misleading**.

<hr />

* This ends our discussion about Ecological correlations. We will continue
with "normal" correlations from now on.

```{r gen-cor, echo=FALSE}
## Generating correlated Normal data
diffr <- function(n, rho, SDx = 1, SDy = 1) {
  meanx <- 3; meany <- 3
  
  x1 <- rnorm(n = n)
  x2 <- rnorm(n = n)
  x3 <- rho*x1 + sqrt(1-rho^2)*x2
  
  x <- meanx + SDx*x1
  y <- meany + SDy*x3
  
  r <- round(cor(x, y), 3)
  
  plot(x, y, xlim = c(0,6), ylim = c(0,6),
       xaxs = "i", yaxs = "i")
}
```

* Question (for the expert in you): you are presented with the following plots
  (below), and asked, due to your expertise, to determine which of the panels
  corresponds to data with a higher correlation coefficient (in absolute value).
  What would be your answer?


```{r plot-cor, echo=FALSE, fig.width=8, fig.height=4, fig.align='center'}
set.seed(1)
par(mai = c(.2, .2, .2, .2), mgp = c(1.5, 0.2, 0),
    tck = -.01, mfrow = c(1,2))
diffr(rho = 0.70, n = 50, SDx = 1.1, SDy = .3)
diffr(rho = 0.70, n = 50, SDx = 1, SDy = 1.3)
```

<br />

### Changing SDs

* The appearance of a scatter diagram depends on the SDs. In fact, both plots
  above were generated with the *same* $r = 0.7$. The difference is that the
  one on the left had $\sigma_X = 1.1$ and $\sigma_Y = 0.3$, while the one on
  the right had $\sigma_X = 1$ and $\sigma_Y = 1.3$.
  However, the left one looks more tightly clustered than the other. It is due
  to the SDs, not $r$ (code in Appendix).
  
* Calculating $r$ involves converting the variables to *standard units*
  where deviations from average are divided by the standard deviation.

* $r$ measures clustering in *relative terms* (relative to the SDs) and not in
  absolute terms.
  
* By the way, how do you feel, from now on, about guessing values of $r$ by
  looking at the scatterplot?

<br />

### Discussion on calculation of the standard deviation

For simplicity, FPP use, unless it was a very small sample, SD instead of sd.
SD is obtained by dividing by $n$ instead of by $n - 1$ (R has the
function `sd` for the latter, and no function for the former).
Anyway, FPP clearly note (later for not confusing people with
technicalities on a book with just a few formulas),
that for small samples one should divide by $n - 1$.

To satisfy people who considers this an important issue,
let us mention that the correct one is the one obtained by
dividing by $n-1$ (because the associated sample variance
is an unbiased estimator of the population variance, while the
other is biased). On the other hand, if the points are
the *population* (not a sample from it),
the correct is the one dividing by $n$.

* By the way, as we are mentioning it, what does it mean that an
  estimator is unbiased?

However, the difference is, on cases say with about 40-50 data points or more,
negligible (if you are using the right kind of data, which *is* what you *should*
be doing), and in the case of Pearson's data (1,078) the *terrible* error
committed by using SD instead of sd is:

```{r}
## FPP version of sample standard deviation
SD <- function(x) sqrt(mean((x-mean(x))^2))

## "Error" of using SD instead of "correct" sd
SD(x) - sd(x)
SD(y) - sd(y)
```

Is that 1/1000 of an inch?

Perhaps, there are far more worse errors one can commit (using data
unsuitable to the statistical procedure comes to mind, or the other way
around, using an unsuitable statistical procedure for the data that, after
all, should drive the analysis), and insisting
too much in technicalities such as this can make statistics rely too
much in mathematical niceties, and too little on substance (just in case,
know that I do not expect that you agree with me in any kind of
remark I make. In fact, many of my remarks are just a consequence of
my lack of certainty about which should be the *correct* answer).

By the way, the maximum likelihood estimator of the variance also divides by
$n$ instead of $n-1$, so it is biased (and used).

We will show both approaches, that lead mostly to the same *final* results,
pointing out when they disagree.

<br />

### The SD line

* The SD line goes through the *point of averages*, passing
  through the points which are an equal number of SDs away from the average,
  for each variable.

* This means that, if you move $n$ $\text{SD}_X$ to the right of $(\bar{X}, \bar{Y})$
  and then $n$ $\text{SD}_Y$ above (or below if $r$ is negative)
  your previous position, you end on a point of the SD line.

* We can find the SD line knowing that it passes throught two points:
\[
\frac{y-y_0}{x-x_0}=\frac{y_1-y_0}{x_1-x_0}
\]
using the points $(\bar{x},\bar{y})$ and $(\bar{x}+\text{SD}_x,\bar{y}+\text{SD}_y)$
\[
\frac{y-\bar{y}}{x-\bar{x}}=\frac{\bar{y}+\text{SD}_y-\bar{y}}{\bar{x}+\text{SD}_x-\bar{x}} \\
\frac{y-\bar{y}}{x-\bar{x}}=\frac{\text{SD}_y}{\text{SD}_x} \\
y-\bar{y}=\frac{\text{SD}_y}{\text{SD}_x} (x-\bar{x})\\
\]

Finally, the SD line using $\text{SD}$ is:

\[
y=\frac{\text{SD}_y}{\text{SD}_x} x+\bar{y}-\frac{\text{SD}_y}{\text{SD}_x}\bar{x}
\]

where $\frac{\text{SD}_y}{\text{SD}_x}$ is the slope and $\bar{y}-\frac{\text{SD}_y}{\text{SD}_x}\bar{x}$
is the intercept.

Of course, the same line is obtained using $\text{sd}$ instead of  $\text{SD}$ (why?):

\[
y=\frac{\text{sd}_y}{\text{sd}_x} x+\bar{y}-\frac{\text{sd}_y}{\text{sd}_x}\bar{x}
\]

where $\frac{\text{sd}_y}{\text{sd}_x}$ is the slope and $\bar{y}-\frac{\text{sd}_y}{\text{sd}_x}\bar{x}$
is the intercept.

```{r SD-line, echo=-c(1:13), fig.width=6, fig.height=6.5, fig.align='center'}
plot(x, y,
     xlim = c(58, 80), ylim = c(58, 80),
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i",
     main = "Pearson's data. SD line",
     xlab = "Father height in inches", ylab = "Son height in inches")
axp <- seq(58, 80, by = 2)
axis(1, at = axp, labels = axp)
axis(2, at = axp, labels = axp)

## Spread of the cloud
abline(v = meanx + c(-3, -2, -1, 1, 2, 3)*sd(x), col = "blue", lty = 3)
abline(h = meany + c(-3, -2, -1, 1, 2, 3)*sd(y), col = "blue", lty = 3)

## Point of averages (center of the cloud)
abline(v = meanx, col = "green")
abline(h = meany, col = "green")

## SD line using equation and FFP SD
abline(a = meany - SD(y)/SD(x)*meanx,
       b = SD(y)/SD(x), col = "blue", lwd = 4)

## SD line using equation and sd
abline(a = meany - sd(y)/sd(x)*meanx,
       b = sd(y)/sd(x), col = "blue", lwd = 4)
```

* Note that we have plotted the SD line twice, using $\text{SD}$ and $\text{sd}$,
obtaining the same result.

* T or F: If you move one SDx to the right of the point of
  averages, would it be your best guess (prediction) of the
  son's height the corresponding point on the SD line?

```{r Points-and-SD-line, echo=FALSE, fig.width=6, fig.height=6.5, fig.align='center'}
plot(x, y,
     xlim = c(58, 80), ylim = c(58, 80), col = "lightgrey",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i",
     main = "Pearson's data. SD line",
     xlab = "Father height in inches", ylab = "Son height in inches")
axp <- seq(58, 80, by = 2)
axis(1, at = axp, labels = axp)
axis(2, at = axp, labels = axp)

## Spread of the cloud
abline(v = meanx + c(-2, 2)*sd(x), col = "blue", lty = 3)
abline(h = meany + c(-2, 2)*sd(y), col = "blue", lty = 3)

## Point of averages (center of the cloud)
abline(v = meanx, col = "green")
abline(h = meany, col = "green")

## SD line using equation and sd
abline(a = meany - sd(y)/sd(x)*meanx,
       b = sd(y)/sd(x), col = "blue", lwd = 4)

fround <- 72
abline(v = fround + c(-.5,.5), lty = 3)
with(subset(father.son, round(fheight,0) == fround),
     points(fheight, sheight, pch = 16,
            col = ifelse(sheight > meany + (fheight - meanx)/sd(x)*sd(y),
                         "darkgreen", "blue")))

fround <- 64
abline(v = fround+c(-.5, .5), lty = 3)
with(subset(father.son, round(fheight,0) == fround),
     points(fheight, sheight, pch = 16,
            col = ifelse(sheight > meany + (fheight - meanx)/sd(x)*sd(y),
                         "darkgreen", "blue")))
```

<br />

### The graph of averages

* If you partition your $x$ axis let's say at every inch, the graph of
  averages is the collection of points where the $x$-coordinate is the
  center of the vertical strip, and the $y$-coordinate the *mean* of all
  the $y$-values contained in that strip. (Of course your selection of
  a different strip width will give a different result, as it happens
  with histograms)

```{r GoA, echo=-c(1:9), fig.width=6, fig.height=6.5, fig.align='center'}
plot(x, y,
     xlim = c(58, 80), ylim = c(58, 80), col = "lightgrey",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i",
     main = "Pearson's data. GoA",
     xlab = "Father height in inches", ylab = "Son height in inches")
axp <- seq(58, 80, by = 2)
axis(1, at = axp, labels = axp)
axis(2, at = axp, labels = axp)

## Point of averages (center of the cloud)
abline(v = meanx, col = "green")
abline(h = meany, col = "green")

## Graph of averages.
sgav <- with(father.son, tapply(sheight, round(fheight, 0), mean))
sgavnum <- with(father.son, tapply(sheight, round(fheight, 0), length))

points(as.numeric(names(sgav)), sgav, col = "red", pch = 16)
text(as.numeric(names(sgav)), sgav, sgavnum, pos = 3)
```

* The GoA is close to a straight line in the middle, but not at the ends.
  Why do you think that happens?

* T or F: If we add to the GoA plot the SD line, would it more or less
  follow the GoA?

```{r SD-GoA, echo=FALSE, fig.width=6, fig.height=6.5, fig.align='center'}
plot(x, y,
     xlim = c(58, 80), ylim = c(58, 80), col = "lightgrey",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i",
     main = "Pearson's data. SD line and GoA",
     xlab = "Father height in inches", ylab = "Son height in inches")
axp <- seq(58, 80, by = 2)
axis(1, at = axp, labels = axp)
axis(2, at = axp, labels = axp)

## Point of averages (center of the cloud)
abline(v=meanx, col="green")
abline(h=meany, col="green")

## SD line using equation and sd
abline(a = meany - sd(y)/sd(x)*meanx,
       b = sd(y)/sd(x), col = "blue", lwd = 4)

## Graph of averages.
sgav <- with(father.son, tapply(sheight, round(fheight, 0), mean))
points(as.numeric(names(sgav)), sgav, col = "red", pch = 16)
```

<br />

### The Regression line and the Regression function

* The *regression line* is a *smoothed version* of the *graph of averages*.
  *If* the graph of averages follows a straight line (as happens on
  a scatterplot that is a *football-shaped* cloud of points),
  that straight line *is* the regression line.

```{r Regression-line, echo=FALSE, fig.width=6, fig.height=6.5, fig.align='center'}
plot(x, y,
     xlim = c(58, 80), ylim = c(58, 80), col = "lightgrey",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i",
     main = "Pearson's data. Regression line and GoA",
     xlab = "Father height in inches", ylab = "Son height in inches")
axp <- seq(58, 80, by = 2)
axis(1, at = axp, labels = axp)
axis(2, at = axp, labels = axp)

## Point of averages (center of the cloud)
abline(v = meanx, col = "green")
abline(h = meany, col = "green")

## Graph of averages.
sgav <- with(father.son, tapply(sheight, round(fheight,0), mean))
points(as.numeric(names(sgav)), sgav, col = "red", pch = 16)

## Regression line
abline(a = meany - r*sd(y)/sd(x)*meanx, b = r*sd(y)/sd(x),
       lwd = 4, col = "red")

fround <- 72
abline(v = fround + c(-.5, .5), lty = 3)
with(subset(father.son, round(fheight, 0) == fround),
     points(fheight, sheight,  pch = 16,
            col = ifelse(sheight > meany + (fheight - meanx)/sd(x)*sd(y)*r,
                         "darkgreen", "blue")))

fround <- 64
abline(v = fround + c(-.5, .5), lty = 3)
with(subset(father.son, round(fheight, 0) == fround),
     points(fheight, sheight, pch = 16,
            col = ifelse(sheight > meany + (fheight - meanx)/sd(x)*sd(y)*r,
                         "darkgreen", "blue")))
```

* In general, we could say that the regression
  *function* (meaning *not limited to a straight line*)
  estimates the **average** value for $y$ corresponding to each value of $x$.
  
* More formally, and implying that we are making probability assumptions
  on $Y$ (that we will see in future notes when turning to a more boring - I mean
  rigorous - stage), the regression function is the conditional
  expectation of $Y$ given $X=x$, or $\text{E}[Y|X=x]$.

* Repeter avec moi: The regression function is the conditional expectation
  of $Y$ given $X=x$.
  
* Informally (I think, but I may be wrong...) we could say that the graph
  of averages is a "discrete" version (as it depends on the width
  of the strip) of the regression *function*
  (linear or non linear), and that if we could make the width of the
  strips go to zero, they should agree (provided we have an infinite number
  of points so we always have points inside the strips as their width vanishes).

* Keep in mind that the regression line may smooth away too much.
  For example, if there is a non-linear association between the two variables,
  the regression line will pass it by.
  (We will see that there are situations where we can transform the data
  and still use linear regression. If not, we need to move to other
  non-linear methodologies).

* Anyway, the graph of averages should *always* work.

* We have related GoA and regression line. Is there any relation between
  SD line and regression line?

* OK, I did not show you the equation of the regression *line* (the
  one that is good for football-shaped clouds of points). We are going
  to give here a version that we will accept without proof.
  Later, we will formally derive another version that relies on calculus,
  and you will then be asked to reconcile both versions. Here it goes:
\[
y = r\frac{\text{SD}_y}{\text{SD}_x}x + 
    \bar{y} - r\frac{\text{SD}_y}{\text{SD}_x}\bar{x}
\]
Or, using $\text{sd}$:
\[
y = r\frac{\text{sd}_y}{\text{sd}_x}x + 
    \bar{y} - r\frac{\text{sd}_y}{\text{sd}_x}\bar{x}
\]

* If you already forgot (I would have), the equation of the SD line follows:

\[
y = \frac{\text{SD}_y}{\text{SD}_x}x + 
    \bar{y} - \frac{\text{SD}_y}{\text{SD}_x}\bar{x}
\]

* Then, do you see any relation between SD line and regression line?

While you think we will put everything in one plot, showing the equation
of the regression line:

```{r Everything, echo=18:19, fig.width=6, fig.height=6.5, fig.align='center'}
plot(x, y,
     xlim = c(58, 80), ylim = c(58, 80), col = "lightgrey",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i",
     main = "Pearson's data. SD line, regression line and GoA",
     xlab = "Father height in inches", ylab = "Son height in inches")
axp <- seq(58, 80, by = 2)
axis(1, at = axp, labels = axp)
axis(2, at = axp, labels = axp)

## Point of averages (center of the cloud)
abline(v = meanx, col = "green")
abline(h = meany, col = "green")

## SD line using equation and sd
abline(a = meany - sd(y)/sd(x)*meanx,
       b = sd(y)/sd(x), col = "blue", lwd = 4)

## Graph of averages.
sgav <- with(father.son, tapply(sheight, round(fheight, 0), mean))
points(as.numeric(names(sgav)), sgav, col = "red", pch = 16)

## Regression line
abline(a = meany - r*sd(y)/sd(x)*meanx, b = r*sd(y)/sd(x),
       lwd = 4, col = "red")

fround <- 72
abline(v = fround + c(-.5, .5), lty = 3)
with(subset(father.son, round(fheight, 0) == fround),
     points(fheight, sheight, pch = 16,
            col = ifelse(sheight > meany + (fheight - meanx)/sd(x)*sd(y)*r,
                         "darkgreen", "blue")))

fround <- 64
abline(v = fround + c(-.5, .5), lty = 3)
with(subset(father.son, round(fheight, 0) == fround),
     points(fheight, sheight, pch = 16,
            col = ifelse(sheight > meany + (fheight - meanx)/sd(x)*sd(y)*r,
                         "darkgreen", "blue")))
```

* The regression line is an *attenuated* version of the SD line. In the SD line,
 when you move 1 SDx to the right, you go 1 SDx above (or below for negative $r$).
 In the regression line, you only go $r$ SDy above (or below).

* Question that is still unanswered from the beginning: we said that
  the correlation coefficient is a measure of linear association,
  or clustering about a line. Which is the line?
  
* Answer:
    - It is the SD line.

<hr />

### Appendix

#### Code for ecological correlations

```{r gen-cor, eval=FALSE}
```
```{r plot-cor, eval=FALSE}
```


#### Code to produce the SD line plot

```{r SD-line, eval=FALSE}
```

#### Code to produce points in relation to the SD line plot

```{r Points-and-SD-line, eval=FALSE}
```

#### Code to produce the Graph of Averages plot

```{r GoA, eval=FALSE}
```

#### Code to produce the Graph of Averages + SD line plot

```{r SD-GoA, eval=FALSE}
```

#### Code to produce the Regression line plot

```{r Regression-line, eval=FALSE}
```

#### Code to produce the plot with everything

```{r Everything, eval=FALSE}
```

### Next

<a href="https://rbertolusso.github.io/posts/LR03-residuals-RMSE">LR03: Residuals and RMSE</a>

### Related

<a href="https://rbertolusso.github.io/posts/intubate-and-stat-functions-in-pipelines">intubate <||> R stat functions in data science pipelines</a>
